---
title: 'RR: Peer assessment 1'
output: pdf_document
---
First the we have to aquire the data:
```{r grab data}
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","project1.zip")
unzip("project1.zip",list = T)
unzip("project1.zip")
activity.data <- read.csv("activity.csv")
```


 Next, we check the structure of the data:
```{r}
 str(activity.data)
```
Now, we transform the date date to YMD format:
```{r}
activity.data$date <- as.Date(activity.data$date, format= "%Y-%m- %d")
```
 #What is mean total number of steps taken per day?
 Calculate the total number of steps taken per day:
 
```{r}
 per_day_total_steps <- aggregate(steps ~ date, activity.data, sum)
 head(per_day_total_steps)
 tail(per_day_total_steps)
```
Make a histogram of the total number of steps taken each day:
```{r} 
lattice::histogram(steps~date, per_day_total_steps, breaks = 6, main = "Histogram of Steps Taken")
```


```{r} 
mean(per_day_total_steps$steps, na.rm = T)
median(per_day_total_steps$steps, na.rm = T) 
```
#What is the average daily activity pattern?
Aggregrate the steps by intervals and plot
```{r} 
interval_steps <- aggregate(activity.data$steps, by= list(interval= activity.data$interval), FUN= mean, na.rm= T)
lattice::xyplot(x ~ interval, data=interval_steps, type= "l", xlab="5 Min Intervals", ylab="Number of Steps", main="Average Daily Activity Pattern")
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r} 
psych::describe(interval_steps)
library(dplyr)
interval_steps %>% filter(x>200.00)
```

#Imputing missing values
Calculate and report the total number of missing values in the dataset:
```{r} 
sum(is.na(activity.data$steps))
```
Devise a strategy for filling in all of the missing values in the dataset and Create a new dataset that is equal to the original dataset but with the missing data filled in:
```{R} 
activity.data_filled <- gam::na.gam.replace(activity.data)
```
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day:
```{r} 
lattice::histogram(steps~date, per_day_total_steps, breaks = 6, main = "Histogram of Steps Taken")
lattice::histogram(steps~date, activity.data_filled, breaks = 6, main = "Histogram of Steps Taken With NA's Replaced")

```
We can see that the new data set is different than the one with NA's, and maybe the gam package is not the best for this:
```{r} 
mean(activity.data_filled$steps)
median(activity.data_filled$steps)
```
#Are there differences in activity patterns between weekdays and weekends?

``` {r} 
weekday_or_end <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) 
        return("weekday") else if (day %in% c("Saturday", "Sunday")) 
        return("weekend") else stop("invalid date")
}

activity.data_filled$date <- as.Date(activity.data_filled$date)
activity.data_filled$day <- sapply(activity.data_filled$date,weekday_or_end)
```

Now, let's make a panel plot containing plots of average number of steps taken on weekdays and weekends.
``` {r} 
avgs <- aggregate(steps ~ interval + day, data = activity.data_filled, mean)
lattice::xyplot(steps ~ interval| as.factor(day), data=avgs, type="l", ylab ="Number of Steps")
```